---
title: "Differential abundance analysis of microbiome count data"
author: 
- name: Fentaw Abegaz
  affiliation: BDA, University of Amsterdam, Netherlands
  email: f.a.yazew@uva.nl
date: 'Compiled: `r format(Sys.Date(), "%d %B, %Y")`'
output: 
    html_document:  
    toc: yes
    #toc_float: yes
    number_section: yes
    highlight: tango
    theme: united
pdf_document:
    toc: yes
vignette: >
  %\VignetteEcoding{UTF-8}
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteIndexEntry{Intro}
  #\usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r options, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(
    warning = FALSE, 
    error = FALSE, 
    message = FALSE)
```
# Load libraries and functions

## Load all the required libraries.
```{r, echo=FALSE, message=FALSE}
library(phyloseq)          
library(DESeq2)             
library(zinbwave)
library(apeglm)    
library(parallel)
library(readr)
library(tibble)
library(dplyr)
library(tidyverse)  
library(ape)
library(GUniFrac)
library(vegan)
#library(devtools)
#install_github("lichen-lab/GMPR")
library(GMPR)
library(ggplot2)
library(RColorBrewer)
library(randomcoloR)
library(pheatmap)
library(reshape2)
library(ggpubr)
library(miceadds) #source.all
library(scales)
# Register BiocParallel Serial Execution 
BiocParallel::register(BiocParallel::SerialParam())
```

## Source all the required .R functions 

These functions are included in the folder ".\\MicroDA\\R". 
```{r, echo=TRUE}

Rsfolder_pathDA <- "..\\R"  ## Replace with the path to your folder

source.all(Rsfolder_pathDA, grepstring="\\.R",  print.source=TRUE )

```   
# Prepare data

## Load the data in phyloseq format 
To maximize the efficiency of the plotting functions provided in this code, it is necessary to group the samples(for example, two groups of 5 samples each can be arranged as: A A A A A B B B B B) and check the taxonomy table headings are in the standard format: Kingdom, Phylum, Class, Order, Family, Genus, and Species.

```{r, echo=TRUE}
# #Forest-Potting soil experimental data
# #Template microbiome datasets used in the paper
# #1. Forest_potting soils experimental data 
# Data_phyloseq <- readRDS("..\\data\\soil_sample_gr2.rds")
# sample_data(Data_phyloseq)$group <- sample_data(Data_phyloseq)$Soil_Type
# 
# #2. N-P starvation experimental data 
# Data_phyloseq <- readRDS(file = "..\\data\\Davar_cleaned_DataHoaglandCompartment.RDS")
# sample_data(Data_phyloseq)$group <- sample_data(Data_phyloseq)$sample_title
# counts <- t(otu_table(Data_phyloseq, 'matrix'))
# group <- sample_data(Data_phyloseq)$Soil_Type 
# taxadata = tax_table(Data_phyloseq)

#
Data_phyloseq <- readRDS("..\\data\\subANDsup_gingival.RDS")
Data_phyloseq
sample_data(Data_phyloseq)$group <- sample_data(Data_phyloseq)$body_subsite
#sort(sample_data(Data_phyloseq)$group)

counts <- t(otu_table(Data_phyloseq, 'matrix'))
group <- sample_data(Data_phyloseq)$body_subsite 
taxadata = tax_table(Data_phyloseq)
   
```

## Groupwise filtering
A two-step filtering strategy is adopted here. In the first step, we keep taxa with a minimum of 2 counts (min_counts=2) in at least 2 replicates or samples per group or experimental condition (min_replicatee=2). These minimum settings can be changed depending on the data. In the second step the retained taxa are classified having group-wise structured zeros or not, depending on the presence/absence of taxa with all zero counts in a group.   

```{r , echo=FALSE}
#Two groups
g_filtered <- group_wise_filer(group=group, counts=counts, min_counts=10, min_replicates=2)
taxa_All_filt <- g_filtered$taxa_All_filt  #all filtered taxa
cat("Total number of filtered taxa = ", ncol(taxa_All_filt),"\n")
 
taxa_NonZeroGroup <- g_filtered$taxa_NonZeroGroup   #filtered taxa in the non-zero-group
cat("Number of taxa in the Non-Zero Group = ", ncol(taxa_NonZeroGroup),"\n")

taxa_ZeroGroup <- g_filtered$taxa_ZeroGroup   #filtered taxa in the zero-group
cat("Number of taxa in the Zero Group = ", ncol(taxa_ZeroGroup),"\n")


```
# Analyze data

## Differential abundance analysis
A differential abundance strategy that combines DESeq2 and DESeq2-ZINBWaVE is implemented to address sparsity (zero-inflation and group-wise structured zeros) and over dispersion in microbiome data. This strategy exploits the penalized likelihood based inference in DESeq2 that helps for testing differential abundance in taxa with group-wise stractutred zeros as well as the power full DESeq2-ZINBWaVE approach, which down-weights inflated-zeros using weights computed based on ZINBWaVE model. DESeq2-ZINBWaVE also has an acceptable FDR control. The methodology is primarily intended for a two-group comparison using likelihood ratio tests. However, it can be extended to the analysis of more than two group or several variables with a carefully defined the design matirx, the full and reduced models for the likelihood ratio test. Either poscounts or GPRM normalization can be used.

The result of the analysis is a table with log-fold changes, p-values, adjusted p-values and taxonomic information for all significant taxa.

```{r , echo=FALSE}


results_D2ZD2 <- DA_DESeq2_ZINBWaVE_DESeq2(taxa_All_filt,taxa_NonZeroGroup, taxa_ZeroGroup, group,  taxadata, reduced=~1, normalization="poscounts") 


#Table of significant taxa
taxa_sig_All_names = union(rownames(results_D2ZD2$sign_taxa_ZeroGroup),
                     rownames(results_D2ZD2$sign_taxa_NonZeroGroup))

results_taxa_ZeroGroup <-results_D2ZD2$results_taxa_ZeroGroup
results_taxa_NonZeroGroup <-results_D2ZD2$results_taxa_NonZeroGroup

results_taxa_ALL <- rbind(results_taxa_ZeroGroup, results_taxa_NonZeroGroup)
results_taxa_ALL

```

## List of significant taxa according to zero-grouping
```{r, echo=FALSE}
cat("List of significant taxa with group-wise structured zeros", "\n")
rownames(results_D2ZD2$sign_taxa_ZeroGroup)

cat("List of significant taxa with no group-wise structured zeros", "\n")
rownames(results_D2ZD2$sign_taxa_NonZeroGroup)

taxa_sig_All_names = union(rownames(results_D2ZD2$sign_taxa_ZeroGroup),
                     rownames(results_D2ZD2$sign_taxa_NonZeroGroup))
cat("List of all significant taxa ",  "\n")
taxa_sig_All_names
```

# Summary plots

The summarpy plots include Heatmap, log-fold change plots and count plots of siginificant taxa.

```{r,fig.width=15, fig.height=10, echo=FALSE}
# table of significant taxa

taxa_sigPWLR = union(rownames(results_D2ZD2$sign_taxa_ZeroGroup),
                     rownames(results_D2ZD2$sign_taxa_NonZeroGroup))
ps_all <- results_D2ZD2$ps_all
results_taxa_ZeroGroup <-results_D2ZD2$sign_taxa_ZeroGroup
results_taxa_NonZeroGroup <-results_D2ZD2$sign_taxa_NonZeroGroup

Da_plots <- plot_results(taxa_sigPWLR, results_taxa_ZeroGroup, results_taxa_NonZeroGroup, ps_all, group=group )
Da_plots
# Da_plots$plot_family
# Da_plots$plot_genus
# Da_plots$plot_counts

```
## Adjusting width and height of individaul plots: heatmap
```{r,fig.width=15, fig.height=10, echo=FALSE}
hmap <- ggpubr::ggarrange(Da_plots$plot_heatmap$gtable,  nrow = 1)
hmap
```
## Adjusting width and height of individaul plots:logfold change plot
```{r,fig.width=10, fig.height=6, echo=FALSE}
Logfold_family <- Da_plots$plot_family 
Logfold_family
```
## Differential abundance analysis for a factor with three levels
```{r}

#load data

#N-P starvation experimental data
Data_phyloseq <- readRDS(file = "C:\\Users\\fenta\\Documents\\OneDrive\\Documents\\2021\\Amsterdam\\MicrobiomeProject\\Davar\\Davar_Fentaw\\data\\Davar_cleaned_DataHoaglandCompartment.RDS") #HoagLand

 # Data_phyloseq <- readRDS(file = "C:\\Users\\fenta\\Documents\\OneDrive\\Documents\\2021\\Amsterdam\\MicrobiomeProject\\Davar\\Davar_Fentaw\\data\\Davar_cleaned_Data.RDS")  
 # 

Data_phyloseq
counts <- t(otu_table(Data_phyloseq, 'matrix'))
group <- sample_data(Data_phyloseq)$sample_title 
taxadata = tax_table(Data_phyloseq)

#Filter data

g_filtered <- group_wise_filer(group=group, counts=counts, min_counts=2, min_replicates=2)
taxa_All_filt <- g_filtered$taxa_All_filt  #all filtered taxa
cat("Total number of filtered taxa = ", dim(taxa_All_filt)[2],"\n")
 
#C vs N 
taxa_NonZeroGroup12 <- g_filtered$taxa_NonZeroGroup12   #filtered taxa in the non-zero-group
cat("Number of taxa in the Non-Zero N-C Group = ", dim(taxa_NonZeroGroup12)[2],"\n")
taxa_ZeroGroup12 <- g_filtered$taxa_ZeroGroup12   #filtered taxa in the zero-group
cat("Number of taxa in the Zero N-C Group = ",dim(taxa_ZeroGroup12)[2],"\n")

#C vs P 
taxa_NonZeroGroup13 <- g_filtered$taxa_NonZeroGroup13   #filtered taxa in the non-zero-group
cat("Number of taxa in the Non-Zero P-C Group = ", dim(taxa_NonZeroGroup13)[2],"\n")
taxa_ZeroGroup13 <- g_filtered$taxa_ZeroGroup13   #filtered taxa in the zero-group
cat("Number of taxa in the Zero P-C Group = ",dim(taxa_ZeroGroup13)[2],"\n")


#Differentila abundance analysis
ps_all <- prune_taxa(colnames(taxa_All_filt), Data_phyloseq)
sample_data(ps_all)$group <- group

#Design matrix
design0 <- model.matrix(~group)
colnames(design0) <- c(colnames(design0)[1],"N_StarvedvsC","P_StarvedvsC")
design00=design0[,2:3]
colnames(design00)=c("N_StarvedvsC","P_StarvedvsC")

sample_data(ps_all)$N_StarvedvsC <- design00[,"N_StarvedvsC"] # factorize for DESeq2
sample_data(ps_all)$P_StarvedvsC <- design00[,"P_StarvedvsC"] # factorize for DESeq2



compare_CvsN <- DA_DESeq2_ZINBWaVE_DESeq2_G3(taxa_All_filt,taxa_NonZeroGroup=taxa_NonZeroGroup12,
                                          taxa_ZeroGroup=taxa_ZeroGroup12, group=group, ps_all=ps_all,
                                       reduced= ~ P_StarvedvsC, tb_tested="N_StarvedvsC",
                                       normalization = "GMPR")

#Table of significant taxa
CN_taxa_sig_All_names = union(rownames(compare_CvsN$sign_taxa_ZeroGroup),
                     rownames(compare_CvsN$sign_taxa_NonZeroGroup))
CN_results_taxa_ZeroGroup <-compare_CvsN$results_taxa_ZeroGroup
CN_results_taxa_NonZeroGroup <-compare_CvsN$results_taxa_NonZeroGroup

CN_results_taxa_ALL <- rbind(CN_results_taxa_ZeroGroup, CN_results_taxa_NonZeroGroup)


compare_CvsP <- DA_DESeq2_ZINBWaVE_DESeq2_G3(taxa_All_filt,taxa_NonZeroGroup=taxa_NonZeroGroup13,
                                          taxa_ZeroGroup=taxa_ZeroGroup13, group=group, ps_all=ps_all,
                                       reduced= ~ N_StarvedvsC, tb_tested="P_StarvedvsC",
                                       normalization = "GMPR")

 #Table of significant taxa
CP_taxa_sig_All_names = union(rownames(compare_CvsP$sign_taxa_ZeroGroup),
                     rownames(compare_CvsP$sign_taxa_NonZeroGroup))

CP_results_taxa_ZeroGroup <-compare_CvsP$results_taxa_ZeroGroup
CP_results_taxa_NonZeroGroup <-compare_CvsP$results_taxa_NonZeroGroup

CP_results_taxa_ALL <- rbind(CP_results_taxa_ZeroGroup, CP_results_taxa_NonZeroGroup)


#List of significant taxa
cat("List of significant taxa in C vs N with group-wise structured zeros", "\n")
rownames(compare_CvsN$sign_taxa_ZeroGroup)

cat("List of significant taxa in C vs N with no group-wise structured zeros", "\n")
 rownames(compare_CvsN$sign_taxa_NonZeroGroup)

cat("List of significant taxa in C vs P with group-wise structured zeros", "\n")
rownames(compare_CvsP$sign_taxa_ZeroGroup)

cat("List of significant taxa in C vs P with no group-wise structured zeros", "\n")
rownames(compare_CvsP$sign_taxa_NonZeroGroup)
```


#Plots
```{r, fig.width=15, fig.height=10, echo=FALSE}

#Plots CvsN
CN_taxa_sigPWLR = union(rownames(compare_CvsN$sign_taxa_ZeroGroup),
                     rownames(compare_CvsN$sign_taxa_NonZeroGroup))
CN_sign_taxa_ZeroGroup <- compare_CvsN$sign_taxa_ZeroGroup
CN_sign_taxa_NonZeroGroup <-compare_CvsN$sign_taxa_NonZeroGroup

CN_Da_plots <- plot_results(taxa_sigPWLR=CN_taxa_sigPWLR, sign_taxa_ZeroGroup=CN_sign_taxa_ZeroGroup, sign_taxa_NonZeroGroup=CN_sign_taxa_NonZeroGroup, ps_all=ps_all, group=group)
suppressWarnings(CN_Da_plots)

#Plots CvsP
CP_taxa_sigPWLR = union(rownames(compare_CvsP$sign_taxa_ZeroGroup),
                     rownames(compare_CvsP$sign_taxa_NonZeroGroup))
CP_sign_taxa_ZeroGroup <-compare_CvsP$sign_taxa_ZeroGroup
CP_sign_taxa_NonZeroGroup <-compare_CvsP$sign_taxa_NonZeroGroup

CP_Da_plots <- plot_results(taxa_sigPWLR=CP_taxa_sigPWLR, sign_taxa_ZeroGroup=CP_sign_taxa_ZeroGroup, sign_taxa_NonZeroGroup=CP_sign_taxa_NonZeroGroup, ps_all=ps_all, group=group) 
suppressWarnings(CP_Da_plots)

```
#Plots
```{r, fig.width=15, fig.height=10, echo=FALSE}
NvsC_logfold <- CN_Da_plots$plot_family 
NvsC_logfold
# #save plot
#  ggsave(filename = "C:\\Users\\fenta\\Documents\\OneDrive\\Documents\\2021\\Amsterdam\\papers\\DifferentialAbundance\\differentialAbundanceComments\\secondAuthorscomments\\DAplots\\NvsC_logfold.png",
#         width=15, height=10)
#  NvsC_logfold
#  dev.off()
```
```{r, fig.width=12, fig.height=8, echo=FALSE}
NvsP_logfold <- CP_Da_plots$plot_family 
NvsP_logfold
# #save plot
#  ggsave(filename = "C:\\Users\\fenta\\Documents\\OneDrive\\Documents\\2021\\Amsterdam\\papers\\DifferentialAbundance\\differentialAbundanceComments\\secondAuthorscomments\\DAplots\\NvsP_logfold.png",
#         width=12, height=8)
#  NvsP_logfold
#  dev.off()
```
