---
title: "Comparing differential abundance (DA) analysis methods for microbiome count data"
author: 
- name: Fentaw Abegaz
  affiliation: BDA, University of Amsterdam, Netherlands
  email: f.a.yazew@uva.nl
date: 'Compiled: `r format(Sys.Date(), "%d %B, %Y")`'
output: 
  pdf_document:
    toc: yes
  html_document:  
    toc: yes
    #toc_float: yes
    number_section: yes
    highlight: tango
    theme: united
vignette: >
  %\VignetteEcoding{UTF-8}
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteIndexEntry{Intro}
  #\usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r options, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(
    warning = FALSE, 
    error = FALSE, 
    message = FALSE)
```


# Load libraries and functions

## Set working directory

```{r setup}
workingDirectory <- getwd()
```

## Load libraries
```{r, echo=TRUE, message=FALSE}
library(pacman)
library(devtools)
library(SparseDOSSA2)
library(pkgmaker)
library(optparse)
library(parallel)
library(stringi)
library(doParallel)
library(edgeR)
library(zinbwave)
library(SummarizedExperiment)
library(dplyr)
library(readr)
library(tibble)
library(BiocParallel)
library(DESeq2)
library(phyloseq)
library(ROCR)
library(plyr)
library(ggplot2)
library(ape)
library(miceadds) #source.all
library(ggpubr)
library(grid)
```

## Load simulation functions 
These .R functions are included in the folders "\\simulation_performance" and "\\sim_fun".

```{r, echo=TRUE}
Rsfolder_pathPerf <- "..\\simulation_performance" 
source.all(Rsfolder_pathPerf, grepstring="\\.R",  print.source=TRUE )

Rsfolder_pathSim <- ".\\sim_func" 
source.all(Rsfolder_pathSim, grepstring="\\.R",  print.source=TRUE )

```
# Simulations

## Simulation parameters
Collect the values set for simulation senarios. The number of microbes (nMicrobes) and read deapth are set from the template dataset. A range of total sample sizes from (ns) in the two groups or experimental conditions and effect sizes or log-fold changes. 

```{r, echo=TRUE}
nMicrobes <- nMicrobes #303
readDepth<- readDepth  #1883

es=es #c(0.5,1,2)   #effect sizes
ns=c(10,20,50,100,200) #total samlple sizes
nIterations = 100  #number of simulations
```


## Calculate performance measures of DA methods
This includes calculating false discovery rates (FDR), power (sensitivity), specificity, AUC, MCC and F1-scores for several differential abundance methods: Deseq2, edgeR and limma-voom, and their ZINBWaVE weighted counterparts Deseq2-ZINBWaVE, edgeR-ZINBWaVE, limma-voom-ZINBWaVE. We also included MaAsLin2. Because many of these techniques treat taxa with group-wise structured zeros differently in differential abundance testing and to ensure a fair comparison of these techniques, taxa without group-wise structured zeros are considered in our simulation-based comparisons. The following function reads the simulated data from the folder 'Input' and save the resulting simulation performance measures in the folder "Output".

```{r, echo=TRUE, results='hide', warning = FALSE}
output_performance <- run_performance_simulations(es, ns, nIterations, nMicrobes,readDepth)
#simulation results saved in the folder "output"

```
## Summarize the various measures of performance by averaging over the simulations
```{r, echo=TRUE}
methodName <- c('DESeq2','edgeR',
                'limmaVOOM',
                'limmaVOOMZinbwave',
                'DESeq2Zinbwave',
                'edgeRZinbwave',
                'MaAsLin2')
 
rResultsAll <- summarizing_performance(nMicrobes,readDepth, nIterations, methodName,ns,es)


rResults1All <- rResultsAll[which(rResultsAll$methodName %in% 
                  c('DESeq2','DESeq2Zinbwave','edgeR','edgeRZinbwave'
                    ,'limmaVOOM','limmaVOOMZinbwave', 'MaAsLin2')),]

rResults1All$nSubjects=factor(rResults1All$nSubjects,
                              levels=c("10","20","50","100","200"))

names(rResults1All)[names(rResults1All)=="nSubjects"] <- "Samples"

```

## Plot FDR and power(sensitivity) across effect and sample sizes.
```{r, fig.width=20, fig.height=10, fig.wide = TRUE, echo=TRUE}

#FDR
rResults1All2=subset(rResults1All,effectSize==0.5)
e1=ggplot(rResults1All2, aes(methodName , FDR, color=Samples)) +    
  geom_point(position = position_jitterdodge()) +
  geom_boxplot(alpha=0.6)+
  theme_bw()+
  theme(text = element_text(size = 12), axis.text = element_text(face="bold")) +
  xlab("Methods") + ylab("FDR") +
  ggtitle("Effect size = 0.5") 


rResults1All5=subset(rResults1All,effectSize==1)
e2=ggplot(rResults1All5, aes(methodName , FDR, color=Samples)) +    
  geom_point(position = position_jitterdodge()) +
  geom_boxplot(alpha=0.6)+
  theme_bw()+
  theme(text = element_text(size = 12), axis.text = element_text(face="bold")) +
  xlab("Methods") + ylab("FDR") +
  ggtitle("Effect size = 1")

rResults1All10=subset(rResults1All,effectSize==2)
e5=ggplot(rResults1All10, aes(methodName , FDR, color=Samples)) +    
  geom_point(position = position_jitterdodge()) +
  geom_boxplot(alpha=0.6)+
  theme_bw()+
  theme(text = element_text(size = 12), axis.text = element_text(face="bold")) +
  xlab("Methods") + ylab("FDR") +
  ggtitle("Effect size = 2")



#Sensitivity
rResults1All2=subset(rResults1All,effectSize==0.5)
s1=ggplot(rResults1All2, aes(methodName , Sensitivity, color=Samples)) +    
  geom_point(position = position_jitterdodge()) +
  geom_boxplot(alpha=0.6)+
  theme_bw()+
  theme(text = element_text(size = 12), axis.text = element_text(face="bold")) +
  xlab("Methods") + ylab("Sensitivity")+
  ggtitle("Effect size = 0.5")

rResults1All5=subset(rResults1All,effectSize==1)
s2=ggplot(rResults1All5, aes(methodName , Sensitivity, color=Samples)) +   
  geom_point(position = position_jitterdodge()) +
  geom_boxplot(alpha=0.6)+
  theme_bw()+
  theme(text = element_text(size = 12), axis.text = element_text(face="bold")) +
  xlab("Methods") + ylab("Sensitivity")+
  ggtitle("Effect size = 1")

rResults1All10=subset(rResults1All,effectSize==2)
s5=ggplot(rResults1All10, aes(methodName , Sensitivity, color=Samples)) +    
  geom_point(position = position_jitterdodge()) +
  geom_boxplot(alpha=0.6)+
  theme_bw()+
  theme(text = element_text(size = 12), axis.text = element_text(face="bold")) +
  xlab("Methods") + ylab("Sensitivity")+
  ggtitle("Effect size = 2")




performanceSim=ggarrange( e1 + rremove("xlab"), s1 + rremove("xlab"), 
                          e2 + rremove("xlab"), s2 + rremove("xlab"),
                          e5 + rremove("xlab"), s5 + rremove("xlab"),
                          common.legend = TRUE,
                          labels = c("A", "D", "B", "E", "C", "F"),  #NULL
                          ncol = 2, nrow =3,
                          align = "hv",
                          font.label = list(size = 10, 
                                            color = "black", face = "bold", 
                                            family = NULL, position = "top")
)
#annotate_figure(fredSim1, bottom = textGrob("Methods", gp = gpar(cex = 1.3)))



performanceSim


#save plot
# ggsave(filename = "C:\\Users\\fenta\\Documents\\OneDrive\\Documents\\2021\\Amsterdam\\papers\\DifferentialAbundance\\differentialAbundanceComments\\secondAuthorscomments\\DAplots\\performanceSim.png",
#        width=18, height=10)
# performanceSim
# dev.off()

```
 Session info
```{r}
#sessionInfo()
```
